---
title: "Shabani_Muller"
author: "Frank Agyemang"
date: "2025-03-30"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading In the data sets
In this chunk of code, we read in the data set and show a table containing the proportion of missing values for each of the chemicals even before they are put into group. The goal of this table is to understand how the proportion of missing values affects our estimates and what statistic would be more effecient for inference. For example, using a mean for this data set is not an efficient the data may be skewed and there are also lot of confounding variables.
```{r}
# Read a specific sheet
tadpole_pest <- read.csv("AltEn_tadpoles_2025_pesticide_summary_030425.csv")
tadpole_det <- read.csv("AltEn_tadpole_detections_with_headers_deidentified.csv")
tadpole_fin <- read.csv("Tadpole_2021_2024_WetWeight_clean.csv")
tadpole_fin$ThiamethoxamD <- as.numeric(tadpole_fin$ThiamethoxamD)
summary(tadpole_fin)

library(dplyr)
tadpole_final <- tadpole_fin %>%
  mutate(
    Neonico = rowSums(tadpole_fin[, c("Clothianidin", "ThiamethoxamD")]),
    Fungi = rowSums(tadpole_fin[, c("Azoxystrobin", "Carbendazim", "Cyproconazole", "Fluoxastrobin", 
                    "Fluxapyroxad", "Ipconazole", "Metconazole", "Propiconazole", 
                    "Pyraclostrobin", "Tebuconazole", "Tetraconazole", "Thiabendazole")]),
    Herbi = rowSums(tadpole_fin[, c("Acetochlor", "Atrazine", "Dithiopyr", "Flufenacet", 
                    "Metolachlor..R.and.S.", "Pendimethalin", "Penoxsulam")]),
    Other = rowSums(tadpole_fin[, c("Cyfluthrin..all.isomers.", "Bifenthrin", "Desthiopro")]),
    Legacy_Agro = rowSums(tadpole_fin[, c("ppDDD", "ppDDE", "ppDDT", "Pentachloroanisole")])
  )


## Removing variables 
tadpole_final <- tadpole_final[, -c(9:36)]
```


```{r}
library(dplyr)
tadpole_final <- tadpole_final %>%
  mutate(ID = c(1:175))
tadpole_final <- tadpole_final %>% select(ID, Collection_date,Site,Specices, everything())


# Proportion of missing values in each column
missing_proportion <- sapply(tadpole_final, function(x) mean(is.na(x)))


# Store it in a data frame
missing_table <- data.frame(
  Variable = names(missing_proportion),
  Missing_Proportion = missing_proportion
)
# View the table
print(missing_table)
```




## Data Cleaning and Exploratory Data Analysis
```{r}
#Treating categorical variables as factors
tadpole_final[, c("Site","Specices","Collection_date","Dev_stage", "ID")] <- lapply(tadpole_final[, c("Site","Specices","Collection_date","Dev_stage","ID")], as.factor)

library(lubridate)
# Convert Date to Date format and extract year
tadpole_final$Collection_date <- year(mdy(tadpole_final$Collection_date))
tadpole_final$Collection_date <- as.factor(tadpole_final$Collection_date)
tadpole_final_freq <- tadpole_final %>%
  count(Specices, Collection_date) %>%
  group_by(Specices, Collection_date) %>%
  mutate(Relative_Freq = n / sum(n))


library(dplyr)

# Define which levels belong to each category
undeveloped_countries <- c("P25","pre-25","p27","P27","26","27-30","28","30")
developing_countries <- c("p31","33","34","35","36","37","38","39")
developed_countries <- c("41")

# Create new category variable
tadpole_final <- tadpole_final %>%
  mutate(Dev_stage = case_when(
    Dev_stage %in% undeveloped_countries ~ "Undeveloped",
    Dev_stage %in% developing_countries ~ "Developing",
    Dev_stage %in% developed_countries ~ "Developed",
    TRUE ~ NA_character_  # or "Other"
  ))

tadpole_final$Dev_stage <- as.factor(tadpole_final$Dev_stage)
tadpole_final <- tadpole_final %>% 
  rename(
    Mass = Mass..g...dry.weight.,
    Moisture = X..Moisture,
    Lipid = X..Lipid
  )
write.csv(tadpole_final, "tadpole_final.csv", row.names = FALSE)
```


# Exploratory Data Analysis

```{r}
library(ggplot2)
# Histogram of Concentration for the various Chemicals
hist(tadpole_final$Neonico)
hist(tadpole_final$Fungi)
hist(tadpole_final$Herbi)
hist(tadpole_final$Other)
hist(tadpole_final$Legacy_Agro)

# Mean Concentration by chemical over Years
conc_neo <- aggregate(Neonico ~ Collection_date, data = tadpole_final, FUN = mean)
conc_fung <- aggregate(Fungi ~ Collection_date, data = tadpole_final, FUN = mean)
conc_herbi <- aggregate(Herbi ~ Collection_date, data = tadpole_final, FUN = mean)
conc_other <- aggregate(Other ~ Collection_date, data = tadpole_final, FUN = mean)
conc_leg <- aggregate(Legacy_Agro ~ Collection_date, data = tadpole_final, FUN = mean)

# Plot 
ggplot(conc_neo, aes(x = Collection_date, y = Neonico)) +
  geom_point(size = 4, color = "black") +
  geom_line(group = 1, color = "black") +
  labs(title = "Mean Plot Of Neonicotinoid Concentration Over the Years", x = "Collection Date", y = "Neonicotinoid Mean Value") +
  theme_minimal()

ggplot(conc_fung, aes(x = Collection_date, y = Fungi)) +
  geom_point(size = 4, color = "black") +
  geom_line(group = 1, color = "black") +
  labs(title = "Mean Plot Of Fungicide Concentration Over the Years", x = "Collection Date", y = "Fungicide Mean Value") +
  theme_minimal()

ggplot(conc_herbi, aes(x = Collection_date, y = Herbi)) +
  geom_point(size = 4, color = "black") +
  geom_line(group = 1, color = "black") +
  labs(title = "Mean Plot Of Herbicide Concentration Over the Years", x = "Collection Date", y = "Herbicide Mean Value") +
  theme_minimal()

ggplot(conc_other, aes(x = Collection_date, y = Other)) +
  geom_point(size = 4, color = "black") +
  geom_line(group = 1, color = "black") +
  labs(title = "Mean Plot Of Other Concentration Over the Years", x = "Collection Date", y = "Other Mean Value") +
  theme_minimal()

ggplot(conc_leg, aes(x = Collection_date, y = Legacy_Agro)) +
  geom_point(size = 4, color = "black") +
  geom_line(group = 1, color = "black") +
  labs(title = "Mean Plot Of Legacy Concentration Over the Years", x = "Collection Date", y = "Legacy Mean Value") +
  theme_minimal()



```


## Within year, concentration levels in different sites

```{r}
conc_neo <- aggregate(Neonico ~ Collection_date +  Site, data = tadpole_final, FUN = mean)
conc_fung <- aggregate(Fungi ~ Collection_date + Site, data = tadpole_final, FUN = mean)
conc_herbi <- aggregate(Herbi ~ Collection_date + Site, data = tadpole_final, FUN = mean)
conc_other <- aggregate(Other ~ Collection_date + Site, data = tadpole_final, FUN = mean)
conc_leg <- aggregate(Legacy_Agro ~ Collection_date + Site, data = tadpole_final, FUN = mean)



conc_neo$Collection_date <- as.numeric(conc_neo$Collection_date)
conc_fung$Collection_date <- as.numeric(conc_fung$Collection_date)
conc_herbi$Collection_date <- as.numeric(conc_herbi$Collection_date)
conc_other$Collection_date <- as.numeric(conc_other$Collection_date)
conc_leg$Collection_date <- as.numeric(conc_leg$Collection_date)

# Plot 1: Mean Neonico concentration per Site across Years
ggplot(conc_neo, aes(x = Collection_date, y = Neonico, color = Site, group = Site)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Mean Neonico Concentration by Site Across Years",
       x = "Collection Date",
       y = "Mean Neonico Concentration") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_x_continuous(breaks = unique(conc_neo$Collection_date))

# Plot 2:
ggplot(conc_fung, aes(x = Collection_date, y = Fungi, color = Site, group = Site)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Mean Fungicide Concentration by Site Across Years",
       x = "Collection Date",
       y = "Mean Fungicide Concentration") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_x_continuous(breaks = unique(conc_fung$Collection_date))

#Plot 3:
ggplot(conc_herbi, aes(x = Collection_date, y = Herbi, color = Site, group = Site)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Mean Herbicide Concentration by Site Across Years",
       x = "Collection Date",
       y = "Mean Herbicide Concentration") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_x_continuous(breaks = unique(conc_herbi$Collection_date))

# Plot 4:
ggplot(conc_other, aes(x = Collection_date, y = Other, color = Site, group = Site)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Mean Other Concentration by Site Across Years",
       x = "Collection Date",
       y = "Mean Other Concentration") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_x_continuous(breaks = unique(conc_other$Collection_date))

#Plot 5:
ggplot(conc_leg, aes(x = Collection_date, y = Legacy_Agro, color = Site, group = Site)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Mean Legacy Agro Concentration by Site Across Years",
       x = "Collection Date",
       y = "Mean Lagacy Concentration") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_x_continuous(breaks = unique(conc_leg$Collection_date))
```


## Regression Models
```{r}
Neonico_model <- glm(Neonico ~ Collection_date + Site + Specices + Mass + Moisture , data = tadpole_final, family = Gamma(link = "log"))
summary(Neonico_model)
plot(Neonico_model)

fungi_model <- glm(Fungi ~ Collection_date + Site + Specices + Mass + Moisture + Lipid, data = tadpole_final, family = Gamma(link = "log"))
summary(fungi_model)
plot(fungi_model)

herbi_model <- glm(Herbi ~ Collection_date + Site + Specices + Mass + Moisture + Lipid, data = tadpole_final, family = Gamma(link = "log"))
summary(herbi_model)
plot(herbi_model)

other_model <- glm(Other ~ Collection_date + Site + Specices + Mass + Moisture + Lipid, data = tadpole_final, family = Gamma(link = "log"))
summary(other_model)
plot(other_model)


```













